### 2022-01-28 chapter6/chapter.tex
same draft

%!TEX root = ../thesis.tex
\chapter{Numerical investigations\label{sec: numerics}}
We have now established multiple methods for approximating fluid queues which suitable for approximating the performance measures for fluid-fluid queues derived in \cite{bo2014}. In this chapter we numerically investigate various aspects of the approximation schemes. Throughout, we compare the QBD-RAP scheme (from Chapters~\ref{sec: conv} and \ref{sec: construction and modelling}), the discontinuous Galerkin method (without limiters or filtering of the solution) (as decribed in Chapter~\ref{ch:galerkin}), and the spatially-coherent uniformisation as described in \cite{bo2013} (which we will refer to a just the \emph{uniformisation scheme}). Recall that the uniformisation scheme is equivalent to a discontinuous Galerkin scheme with a constant basis function in each cell, which is also equivalent to the finite-volume method with an upwind flux. Like the QBD-RAP scheme, due to its stochastic interpretation, the uniformisation scheme is positivity preserving.

Where appropriate, we also compare the above methods with the discontinuous Galerkin method with a slope-limiter implemented, which avoids oscillations and negative solutions. However, recall that, in the context of approximating operators of fluid-fluid queues with these schemes, slope-limiting is not possible, and even when slope limiting is possible, the resultant approximation is at best linear around discontinuities.

The main focus of our numerical experiments is on the convergence properties as the number of basis functions is increased. For discontinuous Galerkin method, this means increasing the number of basis functions used to approximate the solution within each cell. For the QBD-RAP scheme, this means increasing the order of the CME distribution used to construct the scheme. For the uniformisation scheme, this means dividing each cell into smaller sub-cells over which we approximate the solution. As such, each approximation scheme leads to approximations which are matrices of the same size (although not necessarily the same number of non-zero components). For example, if we use 3 basis functions in the discontinuous Galerkin method, we approximate the solution by a quadratic on each cell; the comparable QBD-RAP scheme is to use a CME of order 3; and the comparable uniformisation scheme is to divide each cell into 3 sub-cells of width \(\Delta/3\).

To keep the content of this chapter contained, we do not investigate the performance of the schemes as the cell sizes are decreased, other than for the uniformisation scheme. As such, for each numerical experiment, we keep the cell size fixed. As part of deriving a discontinuous Galerkin scheme, one needs to choose the \emph{numerical flux} which is used to approximate the transition of density from one cell to the next and link all the cells together. We investigate schemes with an upwind flux only.
We do not investigate the stability of the schemes with respect to the step-size of the time-integration or time-integration scheme (where required). Instead, we fix the time-integration step size for each numerical experiment at a suitably small value so as to obey a certain stability criterion (a CFL-like condition, REFs, appendix?). For schemes which require us to integrate over time we always implement the strong stability preserving Runge-Kutta method of order 4 with 5 stages (see appendix and REF). This scheme claims to introduce no more oscillations into the solution as we integrate over time (REFs). Where a slope limiter is implemented, we always implement the \emph{Generalised MUSCL} limiter (REF and Appendix). We do not investigate filtering for the DG scheme (REFs).

The performance of the discontinuous Galerkin method has been well-studied (REFs), and it is well-known that the discontinuous Galerkin method performs remarkably well on problems with smooth solutions (REF). Here, we mostly focus on investigating the numerical performance of the methods on problems with non-smooth solutions, the purpose of this is to emphasise the positivity-preserving properties of the QBD-RAP method. In the stochastic modelling community it is very common to have problems with discontinuous solutions, such as a non-smooth initial conditions (REFs). Even if a fluid-queue is initialised with a smooth initial density, the boundary dynamics may induce transient discontinuities into the problem. There is a very specific set of conditions on the initial density and point masses which must hold for the distribution of the fluid queue to remain smooth as it evolves over time (REFs).Limiting distributions of fluid queues are smooth, however. Further, it is possible that discontinuities are present in the performance measures of fluid-fluid queues.

As we shall see, if the solution is smooth, then the Galerkin method is highly-effective and displays rapid convergence to the solution as the number polynomial basis functions is increased. However, when the solution is non-smooth oscillations and negative approximations may occur when using the discontinuous Galerkin method, leading to non-sensical solutions. In these cases, the QBD-RAP approximation performs relatively well compared to the other positivity preserving schemes investigated.

@@

\section{Function approximation/reconstruction}
We start our investigation by looking at how well the methods perform at approximating an initial condition. For the discontinuous Galerkin method, this means projecting the solution on to the set of polynomial functions which define the scheme, then reconstructing the solution. For the spatially-coherent uniformisation scheme, this means looking at the sub-cell averages. For the QBD-RAP scheme, this means computing the initial vector for the approximation, then reconstructing the solutions as described in Section~\ref{sec: closing}. First we investigate three closing vectors we can use for the reconstruction for the QBD-RAP, from which we find that the closing operator in (\ref{eqn:5462}) performs reasonably compared to the others, and we are able to theoretically prove that it converges. Therefore we use this reconstruction method throughout the rest of the chapter.

For this section we consider approximating the initial condition over a single cell of width \(\Delta = 1\). To numerically evaluate integrals arising in the approximation step (inner products and cell averages) we use a simple trapezoidal rule with 10,001 function evaluations. Similarly, we use 10,001 function evaluations to approximate \(L^2\) norms, and also as a finite set of points over which to compute the KS statistics.

First we briefly investigate the possible choices of closing operators as introduced in Section~\ref{sec: closing}.

\subsection{QBD-RAP closing operators}
The three closing operators we investigate are the following. The unnormalised closing vector in Equation~(\ref{eqn: alal}). A normalised version of the closing operator in Equation~(\ref{eqn:546}), with the normalised version given by (\ref{eqn: nonlin}). We refer to the closing vector in (\ref{eqn: nonlin}) as the \emph{naive normalised} closing vector. Technically we have not yet proved that the \emph{naive normalised} closing vector leads to a convergent approximation scheme, though it clearly appears to converge. The third closing vector we investigate is that in Equation~(\ref{eqn:5462}).

\begin{example}First consider approximating the initial condition with density function \(1(x<0.5)\). In the left-hand side of Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} we plot the Kolmogorov-Smirnov error between the true CDF and the CDF constructed via the QBD-RAP approximation. Interestingly, in this case, the unnormalised error outperforms the other two reconstructions at low orders. It just so happens that, in this case, the error due to truncation of the unnormalised reconstruction destructively interferes with other errors in the reconstruction to cause the error to be lower at low orders for the unnormalised scheme. Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} also shows that the naive normalised and normalised reconstructions perform very similarly -- this is the case throughout much of this subsection. The difference between the naive normalised and normalised reconstructions is how they treat mass in the tail of a matrix exponential distribution (from \(2\Delta\) onward). Intuitively, there should be very little mass in the tail of the distribution as we are using concentrated matrix exponential distributions in the reconstruction.

If we instead look at the \(L^2\) error between the true PDF and the reconstructions in Figure~\ref{fig: fun 2 ks error qbdrap closing vecs}, then we see that the unnormalised reconstruction performs the poorest. Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} also suggests naive normalised and normalised reconstruction are very similar. 
\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun2/ks_error_formatted.pdf}%
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun2/l2_pdf_error_formatted.pdf}
	\caption{KS error between the true CDF, \(2x1(x<0.5)+1(x\geq 0.5)\), and the approximations (left) and \(L^2\) error between the true PDF, \(2\times 1(x<0.5)\), and the approximations (right) for the three closing vectors considered; unnormalised (blue solid line with crosses), naive normalised (orange dashed line with diamonds) and normalised (green dash-dotted line with circles). The naive normalised (orange) an normalised closing vectors are coincident.}
	\label{fig: fun 2 ks error qbdrap closing vecs}
\end{figure}
\end{example}

\begin{example}Now consider approximating the initial density \(f(x)=1\). Observing Figure~\ref{fig: fun 4 ks error qbdrap closing vecs} of the KS error, and \(L^2\) error between the true initial distribution and the reconstructions, we now see that the normalised reconstructions outperform the unnormalised reconstruction. This suggests that, in this case, the `folding' of closing operator about \(\Delta\) has greatly increased the ability of the reconstruction to approximate this initial distribution. 
\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/ks_error_formatted.pdf}%
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/l2_pdf_error_formatted.pdf}
	\caption{KS error between the true CDF, \(x\), and the approximation (left) and \(L^2\) error between the true PDF and the approximation (right) for the three closing vectors considered; unnormalised (blue solid line with crosses), naive normalised (orange dashed line with diamonds) and normalised (green dash-dotted line with circles). The naive normalised (orange) an normalised closing vectors are coincident.}
	\label{fig: fun 4 ks error qbdrap closing vecs}
\end{figure}

Some insight is gained by looking at Figure~\ref{fig: pdf reconstructed} where we plot the reconstructed PDFs for the unnormalised and normalised closing operators for order 1, 3, 5 and 7, as well as the true PDF. Observing Figure~\ref{fig: pdf reconstructed} notice that the unnormalised reconstruction fails to capture the density at the left of each of the plots and for all orders. This feature is due to a significant amount of mass being lost due to the truncation. In comparison the reconstruction with the normalised closing operator is much better in this region due to the `folding' around \(\Delta\) in the construction of the closing operator. In the reconstruction we also need to allocate the initial condition to a phase with positive rate, or negative rate, so that we can use the appropriate approximation and reconstruction method as discussed in Section~\ref{sec: closing}. Here we suppose that the phase is positive, so the `folding' around \(\Delta\) in the normalised closing operators appears appears at the left of the plots. In general, reconstructions via the unnormalised closing operator typically underestimate the value of the function in this region.

Figure~\ref{fig: pdf reconstructed} also show that both closing operators do not approximate the initial condition well at the right-hand side of the interval. This is a common feature, and drawback, of the QBD-RAP method. Perhaps there is a different reconstruction method or yet another closing operator which could alleviate this issue. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{chapter6/figs/qbdrap_closing_vec/fun4/pdfs_formatted.pdf}
	\caption{Reconstructed PDFs using the unnormalised closing operator (blue solid line), normalised closing operator (green dash-dotted line), for various orders, and the true PDF which is \(f(x)=1\) (orange dashed line).}
	\label{fig: pdf reconstructed}
\end{figure} 
\end{example}

\begin{example}Next we consider approximating the initial distribution with density \(-6x^2+6x\). Observing the left-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs}, which plots the KS error against the order of the reconstruction for the three closing operators, we once again see that the reconstruction using the unnormalised closing operator performs the worst, while the performance of the two normalised reconstructions is indistinguishable. However, if we instead look at the right-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs}, which plots the \(L^2\) error between the reconstructed PDF and the true PDF, we see that with this metric, then unnormalised closing operator performs better than the two normalised ones for orders 5 and above. 
\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/ks_error_formatted.pdf}%
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/l2_pdf_error_formatted.pdf}
	\caption{KS error between the true CDF, \(-2x^3+3x^2\), and the approximation (left) and \(L^2\) error between the true PDF \(-6x^2+6x\) and the approximation (right), for the three closing vectors considered; unnormalised (blue solid line with crosses), naive normalised (orange dashed line with diamonds) and normalised (green dash-dotted line with circles). The naive normalised (orange) an normalised closing vectors are coincident.}
	\label{fig: fun 6 ks error qbdrap closing vecs}
\end{figure}

The fact that the unnormalised closing operator outperforms the two normalised ones can be explain, once again, by the fact that, for the normalised operators we have `folded' the ME density function back on itself. In Figure~\ref{fig: pdf reconstructed quadratic} we plot the unnormalised and normalised reconstructions along with the true PDF, \(-6x^2+6x\). In Figure~\ref{fig: pdf reconstructed quadratic} we observe that, at the left of the plot, the normalised reconstructions over estimate the density function in this region, where as the unnormalised reconstruction looks to be doing better. Recall that the `folding' in the normalised closing operator manifest itself as extra mass at the left of these plots compared to the unnormalised reconstructions. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{chapter6/figs/qbdrap_closing_vec/fun6/pdfs_formatted.pdf}
	\caption{Reconstructed PDFs using the unnormalised closing operator (blue solid line), normalised closing operator (green dash-dotted line), for various orders, and the true PDF which is \(f(x)=-6x^2+6x\) (orange dashed line).}
	\label{fig: pdf reconstructed quadratic}
\end{figure} 
\end{example}

\begin{example}Lastly, we consider approximating the initial distribution with PDF \(3e^{-3x}/(1-e^{-3})\). This density function is at a maximum at the left of the region. Considering what we have learnt so far about the unnormalised operator underestimating in this region, we expect that the unnormalised closing operator will perform relatively poorly in this case. Indeed, observing Figures~\ref{fig: fun 7 ks error qbdrap closing vecs} we see that the normalised reconstructions perform relatively well compared to the unnormalised reconstruction as measured by both error metrics (KS statistic and \(L^2\) norm on the PDF). If we observed plots of the PDFs (omitted), we would once again see that this is due to the loss of mass at the left-hand side of the region due to the truncation. 
\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/ks_error_formatted.pdf}%
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/l2_pdf_error_formatted.pdf}
	\caption{KS error between the true CDF, \((1-e^{-3x})/(1-e^{-3})\), and the approximation (left) and \(L^2\) error between the true PDF and the approximation for the three closing vectors considered; unnormalised (blue solid line with crosses), naive normalised (orange dashed line with diamonds) and normalised (green dash-dotted line with circles). The naive normalised (orange) an normalised closing vectors are coincident.}
	\label{fig: fun 7 ks error qbdrap closing vecs}
\end{figure}
\end{example}

@@

\subsection{Comparison of methods}
Here we compare the ability of the QBD-RAP, uniformisation, and DG methods to reconstruct initial conditions.

\begin{example}First we consider the initial condition with CDF \(1(x\geq 0.5)\), that is, a point mass at 0.5. This distribution does not have a PDF so we can compare the CDFs only. In Figure~\ref{fig: fun 1 comp} we plot the KS metric (left) and \(L^1\) metric (right) between the true CDF and the reconstructed approximations. Observing the KS metric, it appears that none of the methods converge and the KS error sits around 0.5. This reflects the fact that convergence in distribution implies point-wise convergence of the CDFs except at points of discontinuity. None of the methods appear to converge at this point. Observing the \(L^1\) error between the true CDF and reconstructed approximation (which is the area between the two CDFs) we now see that the methods appear to show the convergent behaviour we expect. Here, the uniformisation method appears to perform the best, while the QBD-RAP method performs the worst. The rate of convergence of the QBD-RAP method appears to be similar to the rate of convergence of the DG scheme.

>>>>

%!TEX root = ../thesis.tex
\chapter{Numerical investigations\label{sec: numerics}}
We have now established multiple methods for approximating fluid queues which are suitable for approximating the performance measures for fluid-fluid queues derived in \cite{bo2014}. In this chapter we numerically investigate various aspects of the approximation schemes. Throughout, we compare the QBD-RAP scheme (from Chapter~\ref{sec: construction and modelling}), the discontinuous Galerkin method (as described in Chapter~\ref{ch:galerkin}), and the spatially-coherent uniformisation as described in \cite{bo2013}, which we will refer to a just the \emph{uniformisation scheme}. %Recall that the uniformisation scheme is equivalent to a discontinuous Galerkin scheme with a constant basis function in each cell, which is also equivalent to the finite-volume method with an upwind flux. 
Recall that, due to their stochastic interpretation, the uniformisation and QBD-RAP schemes are positivity preserving.

Where possible, we implement the discontinuous Galerkin method with and without a slope-limiter implemented \citep{c99}, \citep[Section~5.6.2]{nodalDGBook} (see also Section~\ref{sec: slope limiting}). The slope limiter avoids oscillations and negative solutions. However, we note that in the context of approximating the operator \(\mathbb\Psi\) for a fluid-fluid queue, it is not obvious how one might apply the concept of slope-limiting, other than to post-process the solution with a limiter or filter. Even when slope limiting is possible, the resultant approximation is at best linear around discontinuities. Moreover, there is a computational cost in applying a limiter. However, high-order accuracy of the DG scheme is maintained away from the discontinuities \citep{c99}, \citep[Section~5.6.2]{nodalDGBook}.

The main focus of our numerical experiments is on the convergence properties as the number of basis functions is increased. For simplicity, the number of basis functions is kept constant across all cells. We refer to the number of basis functions on a cell as the \emph{dimension}. For discontinuous Galerkin method, increasing the dimension of the scheme means increasing the number of polynomial basis functions used to approximate the solution within each cell. E.g.~if we use 3 basis functions in the discontinuous Galerkin method, we approximate the solution by a quadratic on each cell. For the QBD-RAP scheme, increasing the dimension of the scheme means increasing the order of the CME distribution used to construct the scheme. To make a comparable equivalent for the uniformisation scheme with a fixed cell size we divide each cell into smaller sub-cells over which we approximate the solution. I.e.~for a dimension \(p\) uniformisation scheme we divide each cell into \(p\) sub-cells. Equivalently, we may think of a dimension \(p\) uniformisation scheme as using \(p\) piecewise constant functions to approximate the solution on each cell. For all schemes (DG, QBD-RAP and uniformisation), if we construct and order \(p\) approximation, there are \(K\) cells, and \(N\) phases then the resulting approximation to the generator \(\mathbb B\) is a matrix of dimension \(pKN\). In this sense each approximation scheme leads to matrices of the same size (although not necessarily the same number of non-zero elements).

To keep the content of this chapter contained, we do not investigate all aspects of the schemes. For each numerical experiment we keep the cell size fixed for the DG and QBD-RAP schemes. As part of deriving a discontinuous Galerkin scheme, one needs to choose the \emph{numerical flux} which is used to approximate the transition of density from one cell to the next \citep{nodalDGBook}. We investigate schemes with an upwind flux only. For schemes which require us to integrate over time we do not investigate the stability of the schemes with respect to the \(t\)-step-size of the time-integration or time-integration scheme itself (where required). Instead, we fix the time-integration step size for each numerical experiment at a suitably small value to obey a certain stability criterion (a CFL-like condition, \citep[Section~4.8]{nodalDGBook}). Moreover, we always implement the strong stability preserving Runge-Kutta method of order 4 with 5 stages \citep{sr2002}, \citep[Section~5.7]{nodalDGBook} (see also Appendix~\ref{sec: time integration}), which claims to introduce no more oscillations into the solution as we integrate over time. Where a slope limiter is implemented, we implement the \emph{Generalised MUSCL} limiter \citep{c99}, \citep[Section~5.6.2]{nodalDGBook} (see also Section~\ref{sec: slope limiting}). We do not investigate filtering for the DG scheme (see \citep[Section~5.6.1]{nodalDGBook} and references therein).

The performance of the discontinuous Galerkin method has been well-studied in some contexts \cite{c99}, \cite[Section~5.5]{nodalDGBook}, and it is well-known that the discontinuous Galerkin method performs remarkably well on problems with smooth solutions. Here, we mostly focus on investigating the numerical performance of the methods on problems with non-smooth solutions, the purpose of this is to emphasise the positivity-preserving properties of the QBD-RAP method. In the stochastic modelling community it is very common to have problems with discontinuities, such as a non-smooth initial conditions. Even if a fluid-queue is initialised with a smooth initial density, the boundary dynamics may induce transient discontinuities or non-smooth behaviour into the problem (see, for example, Section~\ref{sec: transient approx}, below). A very specific set of conditions must hold for the initial density and point masses for the distribution to remain continuous as it evolves over time \citep{bo2014,bop2020}. Limiting distributions of fluid queues are smooth, however. Further, it is possible that discontinuities are present in the performance measures of fluid-fluid queues (see, for example, Section~\ref{sec: return approx}).

As we shall see, if the solution is smooth, then the Galerkin method is highly-effective and displays rapid convergence to the solution as the number of polynomial basis functions is increased. However, when the solution is non-smooth oscillations and negative approximations may occur when using the discontinuous Galerkin method, leading to nonsense solutions. In these cases, the QBD-RAP approximation performs relatively well compared to the other positivity preserving schemes investigated.

The structure of this chapter is as follows. In Section~\ref{sec: recon num} we compute approximations to various initial conditions for the different methods and observe their performance at approximating the initial condition, which allows us to instrument the performance of the reconstruction methods without considering a specific model or any dynamics of the problem. In Section~\ref{sec: wave num} we investigate a simple travelling wave problem with various initial conditions. For this problem the dynamics are deterministic, which allows us to instrument the ability of the schemes to approximate the flow of probability across cells without any stochastic dynamics. Next, we investigate a simple fluid queue with two phases. Section~\ref{sec:stat} investigates the ability of the methods to approximate the stationary distribution of the model, Section~\ref{sec: transient approx} investigates the ability of the schemes to approximate the transient distribution of the fluid queue for two initial conditions, and Section~\ref{sec: return approx} investigates the ability of the methods to approximate the first hitting time of the fluid level on the boundary of the interval \([0,1]\). Section~\ref{sec: ffq num} approximates the distribution of \(\{(X(t),\varphi(t))\}\) at the time at which \(\{Y(t)\}\) first returns to \(0\) for two simple fluid-fluid queue models.

@@

\section{Function approximation/reconstruction}\label{sec: recon num}
We start our investigation by looking at how well the methods perform at approximating an initial condition. By approximating the initial condition only, we aim to instrument the performance of the approximation schemes without any dynamics.

For the discontinuous Galerkin method, this we project the initial condition on to the set of polynomial functions which define the scheme, for the spatially-coherent uniformisation scheme, we look at the sub-cell averages of the initial condition, and for the QBD-RAP scheme, we compute the initial vector for the approximation, then reconstructing the solutions as described in Sections~\ref{sec: initial conditions} and \ref{sec: closing}. For the purposes of approximation and reconstruction for the QBD-RAP scheme we must orientate the initial condition in a positive or negative direction; here, we suppose that the initial conditions belong to a positive phase. First we investigate three closing vectors we can use for the reconstruction for the QBD-RAP, from which we find that the closing operator in (\ref{eqn:5462}). From the investigation, we decide to use the closing vector (\ref{eqn:5462}) throughout the rest of the chapter.

For this investigation we consider approximating initial conditions over a single cell of width \(\Delta = 1\). To numerically evaluate integrals arising in the approximation step (inner products and cell averages) we use a trapezoidal rule with 10,001 function evaluations. Similarly, we use 10,001 function evaluations to approximate \(L^p\) norms, and also as a finite set of points over which to compute the KS statistics.

% First we briefly investigate the possible choices of closing operators as introduced in Section~\ref{sec: closing}.

\subsection{QBD-RAP closing operators}
The three closing operators we investigate are the following. The unnormalised closing vector in Equation~(\ref{eqn: alal}). A normalised version of the closing operator in Equation~(\ref{eqn:546}), with the normalised version given by (\ref{eqn: nonlin}). We refer to the closing vector in (\ref{eqn: nonlin}) as the \emph{naive normalised} closing vector. Technically we have not yet proved that the \emph{naive normalised} closing vector leads to a convergent approximation scheme, though it clearly appears to converge. The third closing vector we investigate is that in Equation~(\ref{eqn:5462}) and we refer to this as the \emph{normalised closing operator}.

\begin{example}First consider approximating the initial condition with density function \(2\times 1(x<0.5)\). In the left-hand side of Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} we plot the Kolmogorov-Smirnov error between the true CDF and the CDF constructed via the QBD-RAP approximation with the three different closing vectors. Interestingly, in this case, the unnormalised error outperforms the other two reconstructions at low orders. It just so happens that, in this case, the error due to truncation of the unnormalised reconstruction destructively interferes with other errors to cause the error to be lower at low orders for the unnormalised scheme. Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} also shows that the naive normalised and normalised reconstructions perform very similarly -- this is the case throughout much of this subsection. The difference between the naive normalised and normalised reconstructions is how they treat mass in the tail of a matrix exponential distribution (from \(2\Delta\) onward). Intuitively, there should be very little mass in the tail of the distribution as we are using concentrated matrix exponential distributions in the reconstruction.

If we instead look at the \(L^2\) error between the true PDF and the reconstructions in Figure~\ref{fig: fun 2 ks error qbdrap closing vecs}, then we see that the unnormalised reconstruction performs the poorest. Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} also suggests the naive normalised and normalised reconstruction are very similar. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun2/ks_error_formatted.pdf}%
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun2/l2_pdf_error_formatted.pdf}
	\caption{KS error between the true CDF, \(2x1(x<0.5)+1(x\geq 0.5)\), and the approximations (left) and \(L^2\) error between the true PDF, \(2\times 1(x<0.5)\), and the approximations (right) for the three closing vectors considered; unnormalised (blue solid line with crosses), naive normalised (orange dashed line with diamonds) and normalised (green dash-dotted line with circles). The naive normalised (orange) an normalised closing vectors are coincident.}
	\label{fig: fun 2 ks error qbdrap closing vecs}
\end{figure}
\exampleFloatBarrier
\end{example}

\begin{example}Now consider approximating the initial density \(f(x)=1\). Observing Figure~\ref{fig: fun 4 ks error qbdrap closing vecs} of the KS error and \(L^2\) error of the PDF we now see that the normalised reconstructions outperform the unnormalised reconstruction. This suggests that, in this case, the `folding' of closing operator about \(\Delta\) has greatly increased the ability of the reconstruction to approximate this initial distribution. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/ks_error_formatted.pdf}%
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/l2_pdf_error_formatted.pdf}
	\caption{KS error between the true CDF, \(x\), and the approximation (left) and \(L^2\) error between the true PDF and the approximation (right) for the three closing vectors considered; unnormalised (blue solid line with crosses), naive normalised (orange dashed line with diamonds) and normalised (green dash-dotted line with circles). The naive normalised (orange) an normalised closing vectors are coincident.}
	\label{fig: fun 4 ks error qbdrap closing vecs}
\end{figure}

Some insight is gained by looking at Figure~\ref{fig: pdf reconstructed} where we plot the reconstructed PDFs for the unnormalised and normalised closing operators for dimension 1, 3, 5 and 7, as well as the true PDF. Observing Figure~\ref{fig: pdf reconstructed} notice that the unnormalised reconstruction fails to capture the density at the left of each of the plots. This feature is due to a significant amount of mass being lost due to the truncation. In comparison, the reconstruction with the normalised closing operator is much better in this region due to the `folding' around \(\Delta\) in the construction of the closing operator. The `folding' in the normalised closing operator manifest itself as more mass at the left of these plots compared to the unnormalised reconstructions. % Recall that, as discussed in Section~\ref{}, for the reconstruction we must to allocate the initial condition to a phase with positive rate, or negative rate, so that we can use the appropriate approximation and reconstruction method as discussed in Section~\ref{sec: closing}. Here we suppose that the phase is positive, so the `folding' around \(\Delta\) in the normalised closing operators appears appears at the left of the plots. In general, reconstructions via the unnormalised closing operator typically underestimate the value of the function in this region.

Figure~\ref{fig: pdf reconstructed} also show that both closing operators do not approximate the initial condition well at the right-hand side of the interval. Perhaps there is a different reconstruction method or yet another closing operator which could alleviate this issue. 
\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth,trim={0cm 1.25cm 0cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/pdfs_formatted.pdf}
	\caption{Reconstructed PDFs using the unnormalised closing operator (blue solid line), normalised closing operator (green dash-dotted line), for various dimensions, and the true PDF which is \(f(x)=1\) (orange dashed line).}
	\label{fig: pdf reconstructed}
\end{figure} 
\exampleFloatBarrier
\end{example}

\begin{example}Next we consider approximating the initial distribution with density \(-6x^2+6x\). Observing the left-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs}, which plots the KS error against the dimension of the reconstruction for the three closing operators, we once again see that the reconstruction using the unnormalised closing operator performs the worst, while the performance of the two normalised reconstructions is indistinguishable. However, if we instead look at the right-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs}, which plots the \(L^2\) error between the reconstructed PDF and the true PDF, then the unnormalised closing operator performs better than the two normalised ones for dimension 5 and above. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/ks_error_formatted.pdf}%
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/l2_pdf_error_formatted.pdf}
	\caption{KS error between the true CDF, \(-2x^3+3x^2\), and the approximation (left) and \(L^2\) error between the true PDF \(-6x^2+6x\) and the approximation (right), for the three closing vectors considered; unnormalised (blue solid line with crosses), naive normalised (orange dashed line with diamonds) and normalised (green dash-dotted line with circles). The naive normalised (orange) an normalised closing vectors are coincident.}
	\label{fig: fun 6 ks error qbdrap closing vecs}
\end{figure}

The fact that the unnormalised closing operator outperforms the two normalised ones can be explained by the `folding' of the normalised operators around \(\Delta\). In Figure~\ref{fig: pdf reconstructed quadratic} we plot the unnormalised and normalised reconstructions along with the true PDF, \(-6x^2+6x\). In Figure~\ref{fig: pdf reconstructed quadratic} at the left of the plots, we observe that the normalised reconstructions over-estimate the density function, whereas the unnormalised reconstruction looks to be doing better. The `folding' in the normalised closing operator manifest itself as more mass at the left of these plots compared to the unnormalised reconstructions. 
\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth,trim={0cm 1.25cm 0cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/pdfs_formatted.pdf}
	\caption{Reconstructed PDFs using the unnormalised closing operator (blue solid line), normalised closing operator (green dash-dotted line), for various orders, and the true PDF which is \(f(x)=-6x^2+6x\) (orange dashed line).} 
	\label{fig: pdf reconstructed quadratic}
\end{figure} 
\exampleFloatBarrier
\end{example}

\begin{example}Lastly, we consider approximating the initial distribution with PDF \(3e^{-3x}/(1-e^{-3})\). This density function is at a maximum at the left of the region. Considering what we have learnt so far about the unnormalised operator underestimating in this region, we expect that the unnormalised closing operator will perform relatively poorly in this case. Indeed, observing Figures~\ref{fig: fun 7 ks error qbdrap closing vecs} we see that this is the case; the normalised reconstructions perform relatively well compared to the unnormalised reconstruction, as measured by both error metrics (KS statistic and \(L^2\) norm on the PDF). If we observed plots of the PDFs (omitted), we would once again see that this is due to the loss of mass at the left-hand side of the region due to the truncation. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/ks_error_formatted.pdf}%
	\includegraphics[width=0.5\textwidth,trim={1.25cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/l2_pdf_error_formatted.pdf}
	\caption{KS error between the true CDF, \((1-e^{-3x})/(1-e^{-3})\), and the approximation (left) and \(L^2\) error between the true PDF and the approximation for the three closing vectors considered; unnormalised (blue solid line with crosses), naive normalised (orange dashed line with diamonds) and normalised (green dash-dotted line with circles). The naive normalised (orange) and normalised closing vectors are coincident.}
	\label{fig: fun 7 ks error qbdrap closing vecs}
\end{figure}
\exampleFloatBarrier
\end{example}

@@

\FloatBarrier
\subsection{Comparison of methods}\label{sec: comp}
Here we compare the ability of the QBD-RAP, uniformisation, and DG methods to reconstruct initial conditions.

\begin{example}First we consider the initial condition with CDF \(1(x\geq 0.5)\), that is, a point mass at 0.5. This distribution does not have a PDF, so we compare the CDFs only. In Figure~\ref{fig: fun 1 comp} we plot the KS metric (left) and \(L^1\) metric (right) between the true CDF and the reconstructed approximations. Observing the KS metric, it appears that none of the methods converge and the KS error sits around 0.5. This reflects the fact that convergence in distribution implies point-wise convergence of the CDFs except, perhaps, at points of discontinuity. None of the methods appear to converge at the discontinuity at \(x=0.5\). Observing the \(L^1\) error between the true CDF and reconstructed approximation (which is the area between the two CDFs) we now see that the methods appear to show the convergent behaviour we expect. Here, the uniformisation method appears to perform the best, while the QBD-RAP method performs the worst. The rate of convergence of the QBD-RAP method appears to be similar to the rate of convergence of the DG scheme.
